---
title: "FinnGen"
author: "Jennifer Collister"
date: "04/10/2024"
output: html_document
---

```{r setup, include=FALSE}

library(yaml)
library(here)
library(data.table)
library(R.utils)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(qqman)
library(ggpubr)
library(janitor)
library(ggrepel)
library(forcats)
library(tictoc)

knitr::opts_chunk$set(echo = TRUE)
# Set the root directory to the project directory (otherwise knitr works in the directory of the Rmd file by default)
knitr::opts_knit$set(root.dir = here::here())

# Load the project config file for filepaths etc
config = read_yaml(here::here("./config.yml"))

# Specify a safe tmp dir
unixtools::set.tempdir(config$tmp)

# Source shared functions
source(config$scripts$shared)


```

# Read in FinnGen summary data

Browse in https://r12.finngen.fi/lof/

Download from [google cloud bucket](https://console.cloud.google.com/storage/browser/finngen-public-data-r12/lof?inv=1&invt=AbkWfA&pageState=(%22StorageObjectListTable%22:(%22f%22:%22%255B%255D%22)))

```{r}

# Downloaded from google cloud bucket
finngen <- data.table::fread("/home/jenniferco/data/FinnGen/R12/lof_data_finngen_R12_lof.txt.gz") 
head(finngen)

finngen <- finngen %>%
  filter(str_detect(toupper(PHENO), "BREAST"))

table(finngen$PHENO)

finngen %>% filter(PHENO == "C3_BREAST_EXALLC") %>% mutate(P = 10^(-LOG10P)) %>% arrange(P) #%>% select(ID, BETA, P)
finngen %>% filter(PHENO == "C3_BREAST_ERPLUS_EXALLC") %>% mutate(P = 10^(-LOG10P)) %>% arrange(P) %>% select(ID, BETA, P)

```

Looks like 

* FinnGen burden tests were conducted using Regenie
* The available phenotypes include subtypes (ERPLUS, ERNEG)
* All the N diagnoses are other disorders

Phenotype of main interest = [C3_BREAST_EXALLC](https://risteys.finregistry.fi/endpoints/C3_BREAST_EXALLC)

This is excluding all other cancers - there isn't a phenotype that doesn't.

```{r}

finn <- finngen %>% 
  filter(PHENO %in% c("C3_BREAST_EXALLC", "C3_BREAST_ERPLUS_EXALLC")) %>%
  separate_wider_delim(ID, delim=".Mask1.", names=c("SYMBOL", "FREQ"), too_many="merge") %>%
  mutate(
    MASK = "PTV",
    z_score = BETA/SE,
    P = 10^(-LOG10P),
    ID = paste0(SYMBOL, "_", MASK)
    ) %>%
  rename(number_chrom = CHROM,
         pos = GENPOS)

head(finn)

table(finn$number_chrom)

```

# Read MWS, BCAC/UKB and AoU/MGC

```{r}

mws <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/mws.csv") %>%
  filter(MASK == "PTV")
bcac <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/bcac.csv") %>%
  filter(MASK == "PTV")
jurgens <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/jurgens.csv") %>%
  filter(MASK == "PTV")

```

# Compare list of genes

```{r}

genes <- finn %>% select(SYMBOL, number_chrom, pos) %>% inner_join(mws %>% filter(MASK=="PTV", FREQ=="0.001") %>% select(SYMBOL, chr = number_chrom, p = pos), by="SYMBOL")

```


# Meta-analysis of MWS/BCAC/UKB/AoU/MGB/FG

```{r}

combine <- mws %>% 
  select(ID, SYMBOL, MASK, number_chrom, pos, P.mws = P, z_score.mws = z_score) %>%
  inner_join(
    bcac %>% 
      select(ID, SYMBOL, MASK, number_chrom, pos, P.bcac = P, z_score.bcac = z_score), 
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) %>%
  inner_join (
    jurgens %>% 
      select(ID, SYMBOL, MASK, number_chrom, pos, P.jurgens = P, z_score.jurgens = z_score), 
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) %>%
  inner_join (
    finn %>%
      filter(PHENO == "C3_BREAST_EXALLC") %>%
      select(ID, SYMBOL, MASK, number_chrom, finn_pos = pos, P.finngen = P, z_score.finngen = z_score),
    by=c("ID", "SYMBOL", "MASK", "number_chrom")
  )

weights <- combine %>% 
  filter(SYMBOL == "CHEK2", MASK == "PTV") %>% 
  mutate(
    b = z_score.bcac / z_score.mws,
    j = z_score.jurgens / z_score.mws,
    f = z_score.finngen / z_score.mws,
    m = 1) %>%
  select(m, b, j, f) %>%
  as.list()

meta_finn <- combine %>% 
  mutate(
    num = (z_score.mws * weights$m) + (z_score.bcac * weights$b) + (z_score.jurgens * weights$j) + (z_score.finngen * weights$f),
    denom = sqrt(weights$m^2 + weights$b^2 + weights$j^2 + weights$f^2),
    z_score.meta = num/denom,
    P.meta = 2*pnorm(q=-abs(z_score.meta)),
    oriented_beta = NA
    ) %>%
  select(-starts_with("weight_"), -num, -denom) %>%
  pivot_longer(cols=c(starts_with("P."), starts_with("z_score.")), names_sep="\\.", names_to=c(".value", "Study"))

head(meta_finn)

meta_ids <- meta_finn %>% pull(ID) %>% unique()

meta_pos <- meta_finn %>% select(number_chrom, pos, finn_pos) %>% distinct() %>% mutate(diff = finn_pos - pos)
head(meta_pos)
ggplot(aes(x=pos, y=finn_pos, color=number_chrom), data=meta_pos) + geom_point()

meta_finn %>% filter(is.na(SYMBOL))

data.table::fwrite(meta_finn, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta_finn.csv")

```

# Manhattan plots

```{r}

masks <- c("PTV")
mask_labels <- c(PTV = "PTV")
datasets <- list("MWS" = "mws", "BCAC.UKB" = "bcac", "AoU.MGB" = "jurgens", "FinnGen" = "finngen", "Meta" = "meta")
colours <- c("MWS" = "#FF0000", "BCAC.UKB" = "#0047AB", "AoU.MGB" = "#097969", "FinnGen" = "#AB4400", "Meta" = "#000068")

tablist <- list(PTV=NULL)

for(mask in masks) {
  
  mask_data <- meta %>% filter(MASK==mask)
  
  pval_list <- list(MWS=NULL, BCAC.UKB=NULL, AoU.MGB=NULL, FinnGen=NULL, Meta=NULL)
  
  n_gene <- mask_data %>% select(SYMBOL) %>% n_distinct()
  sig <- 0.05/n_gene
  
  min_p <- mask_data %>% pull(P) %>% min()
  ylim <- abs(floor(log10(min_p))) + 2
  
  gene_list <- NULL
  
  meta_genes <- mask_data %>% 
    filter(Study=="meta", P < sig) %>% 
    arrange(P) %>%
    pull(SYMBOL)
  
  for (dataset in names(datasets)) {
    print(paste0(mask, ", ", dataset))
    
    study_data <- mask_data %>% filter(Study == datasets[[dataset]])
    
    regenie_man_plot <- manplot_func(
      sum_stats = study_data %>% mutate(number_chrom=as.character(number_chrom)), 
      trait_name=paste0(mask, " variants"),
      gene_based=TRUE,
      sig = sig,
      ylim = ylim,
      highlight_color = colours[dataset],
      gene_list = gene_list,
      title_text=paste0(mask_labels[mask], " variants in ", dataset, " with sig=", formatC(sig, format = "e", digits = 2))
      )
    
    tiff(paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/figs/genebased_meta/finngen/", dataset, "_", mask, "_Manhattan.tiff"), 
         units="in", width=8, height=7, res=300)
    print(regenie_man_plot)
    dev.off()
    
    pval_list[[dataset]] <- study_data %>% 
      filter(SYMBOL %in% meta_genes) %>% 
      select(SYMBOL, P) %>%
      arrange(P) %>%
      mutate("{dataset}" := formatC(P, format = "e", digits = 2)) %>%
      select(-P)
  }
  tablist[[mask]] <- pval_list %>% purrr::reduce(left_join, by = "SYMBOL")
}

```

