---
title: "BCAC_Annotation"
author: "Jennifer Collister"
date: "22/08/2024"
output: html_document
---

```{r setup, include=FALSE}

library(yaml)
library(here)
library(data.table)
library(R.utils)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(qqman)
library(ggpubr)
library(janitor)
library(ggrepel)
library(forcats)
library(DBI)
library(duckdb)
library(dbplyr)
library(tictoc)
library(jsonlite)
library(httr2)
library(openxlsx)
library(purrr)

knitr::opts_chunk$set(echo = TRUE)
# Set the root directory to the project directory (otherwise knitr works in the directory of the Rmd file by default)
knitr::opts_knit$set(root.dir = here::here())

# Load the project config file for filepaths etc
config = read_yaml(here::here("./config.yml"))

# Specify a safe tmp dir
unixtools::set.tempdir(config$tmp)

# Source shared functions
source(config$scripts$shared)

```

This markdown should be run after BCAC_Annotation.Rmd which adds Helix annotations, creates BCAC masks and provides the code to run gene-based tests in Regenie.

# Read Regenie outputs

Combine across masks and calculate z-scores.

The tests are two-tailed, so divide the P value by 2. The normal distribution is symmetric, so the z score is ±.

```{r, eval=FALSE}

regenie <- vector("list", 3)
i <- 1
for(file in c("rare", "cadd", "helix")) {
  filename = paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/regenie/step2_gene/binary/breast_cancer_",
                    file, "_breast_cancer.regenie")
  data <- data.table::fread(filename)
  
  # Rearrange
  data <- data %>% 
    separate_wider_delim(cols=ID, delim=".", names=c("SYMBOL", "MASK", "FREQ"), too_many="merge", too_few="align_start") %>%
    mutate(oriented_beta = BETA,
           P = 10^(-LOG10P),
           number_chrom=as.numeric(CHROM),
           pos = GENPOS)
  
  if(file == "rare") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV", MASK == "M2" ~ "Rare", TRUE ~ "Uhoh"))
    } else if(file == "cadd") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.CADD", MASK == "M2" ~ "CADD", TRUE ~ "Uhoh"))
    } else if(file == "helix") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.Helix", MASK == "M2" ~ "Helix", TRUE ~ "Uhoh"))
    }
  data <- data %>% mutate(ID = paste0(SYMBOL, "_", MASK, "_", FREQ, "_", TEST))
  regenie[[i]] <- data
  i <- i + 1
}

mws <- dplyr::bind_rows(regenie) %>% 
  filter(
    FREQ == "all",
    TEST == "ADD",
    MASK %in% c("PTV", "Rare", "PTV.CADD", "PTV.Helix")
  ) %>%
  mutate(
    ID = paste0(SYMBOL, "_", MASK),
    z_score = BETA/SE,
    MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix"))
    )
  
rm(regenie)

# data.table::fwrite(mws, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/mws.csv")

```

```{r}

mws <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/mws.csv")

```

# Read in the BCAC gene-level outputs

```{r, eval=FALSE}

bcac_path <- "/home/jenniferco/data/Wilcox2023"
files <- list(PTV = "GCST90267995.tsv", Rare = "GCST90267996.tsv", PTV.CADD = "GCST90267997.tsv", PTV.Helix = "GCST90267998.tsv")

bcac_sep  <- vector("list", 3)
i <- 1
for(file in names(files)){
  data <- data.table::fread(file.path(bcac_path, files[[file]])) %>% 
    janitor::clean_names() %>%
    rename(SYMBOL = gene,
           P = p_value) %>%
    mutate(MASK = file)
  
  bcac_sep[[i]] <- data
  i <- i + 1
}
bcac <- dplyr::bind_rows(bcac_sep) %>% 
  inner_join(mws %>% select(number_chrom, pos, SYMBOL) %>% distinct(), by="SYMBOL") %>%
  mutate(ID = paste0(SYMBOL, "_", MASK),
         oriented_beta = NA,
         MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix")))

head(bcac)

data.table::fwrite(bcac, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/bcac.csv")

```

```{r}

bcac <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/bcac.csv")

```

# All of Us

The [Jurgens2024 ](https://www.nature.com/articles/s41588-024-01894-5) meta-analysis provides summary stats for All of Us and Mass General Brigham Biobank which we can use instead of the All by All - they contain slightly more AoU breast cancer cases.

They have the following masks:

> * hclof_noflag_POPMAX0.001 == LOF mask with MAF<0.1%
* hclof_noflag_missense0.8_POPMAX0.001 == LOF + missense0.8 mask with MAF<0.1%
* hclof_noflag_missense0.8_POPMAX0.001 == LOF + missense0.5 mask with MAF<0.1%
* hclof_noflag_missense0.5_POPMAX0.00001 == LOF + missense0.5 mask with MAF<0.001%
* missense0.5_POPMAX0.00001 == missense0.5 mask with MAF<0.001%
* missense0.2_POPMAX0.00001 == missense0.2 mask with MAF<0.001%
* hclof_noflag_POPMAX0.01 == LOF mask with MAF<1%
* hclof_noflag_missense0.8_POPMAX0.01 == LOF + missense0.8 mask with MAF<1%
* hclof_noflag_missense0.5_POPMAX0.01 == LOF + missense0.5 mask with MAF<1%

We will map:

* PTV = hclof_noflag_POPMAX0.001
* Rare missense = missense0.2_POPMAX0.00001
* CADD and Helix = either hclof_noflag_missense0.8_POPMAX0.001 or hclof_noflag_missense0.5_POPMAX0.001

```{r, eval=FALSE}

# dbname <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/anno_full.db"
# con <- DBI::dbConnect(duckdb::duckdb(), dbname)
# 
# colnames <- tbl(con, "annotation") %>% head() %>% collect()
# 
# genes <- tbl(con, "annotation") %>%
#   filter(Feature_type == "Transcript",
#          (is.na(MANE_SELECT) & CANONICAL == "YES") | !is.na(MANE_SELECT),
#          BIOTYPE == "protein_coding"
#          ) %>%
#   select(Gene, SYMBOL) %>%
#   distinct() %>%
#   collect()
# 
# DBI::dbDisconnect(con, shutdown=TRUE)
# 
# Jurgens_AoU_MGB <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/JurgensWang2024_all_files/All_Mixed_raw/AoU_250k_MGB_53k__10CarriersPerStudy_20Overall_cauchy_harmonizedEstimates.tsv.gz") %>%
#   separate_wider_delim(total, delim="__", names=c("Pheno", "Gene")) %>%
#   filter(Pheno=="Breast_cancer") %>%
#   inner_join(genes, by="Gene")
# 
# data.table::fwrite(Jurgens_AoU_MGB, file="/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/AoU_MGB_All_breast_cancer.tsv")

Jurgens_AoU_MGB <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/AoU_MGB_All_breast_cancer.tsv") %>% 
  select(
    Gene, SYMBOL, cases, controls, P_cauchy, P_cauchy_v2,
    ends_with("hclof_noflag_POPMAX0.001"), # Meta with PTV
    ends_with("missense0.2_POPMAX0.00001"), # Meta with rare missense
    ends_with("hclof_noflag_missense0.8_POPMAX0.001"), ends_with("hclof_noflag_missense0.5_POPMAX0.001") # Meta with PTV_CADD and PTV_Helix
    ) %>%
  rename(popcases = cases, popcontrols=controls) %>%
  pivot_longer(cols=contains("POPMAX"), names_to=c(".value", "raw_mask"),  names_sep="__") %>%
  inner_join(mws %>% select(number_chrom, pos, SYMBOL) %>% distinct(), by="SYMBOL") %>%
  filter(!is.na(P_Meta)) %>%
  mutate(
    P = P_Meta,
    old_z = qnorm(p=P_Meta/2, lower.tail=FALSE),
    z_score = Est_Meta/Est.SE_Meta,
    oriented_beta = Est_Meta,
    ID = SYMBOL
    ) %>%
  filter(!is.na(z_score))

jurgens <- Jurgens_AoU_MGB %>%
  filter(raw_mask != "hclof_noflag_missense0.5_POPMAX0.001") %>%
  mutate(MASK = dplyr::case_when(
    raw_mask == "hclof_noflag_POPMAX0.001" ~ "PTV",
    raw_mask == "missense0.2_POPMAX0.00001" ~ "Rare",
    raw_mask == "hclof_noflag_missense0.8_POPMAX0.001" ~ "PTV.CADD"
    )
  ) %>%
  rbind(
    Jurgens_AoU_MGB %>% 
      filter(raw_mask == "hclof_noflag_missense0.8_POPMAX0.001") %>% 
      mutate(MASK = "PTV.Helix")
  ) %>%
  mutate(
    ID = paste0(SYMBOL, "_", MASK),
    MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix"))
    )

data.table::fwrite(jurgens, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/jurgens.csv")

```

```{r}

jurgens <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/jurgens.csv")

```

# Meta-analysis of MWS/BCAC/UKB/AoU/MGB

```{r, eval=FALSE}

combine <- mws %>% 
  select(ID, SYMBOL, MASK, number_chrom, pos, P.mws = P, z_score.mws = z_score, oriented_beta.mws = oriented_beta) %>%
  inner_join(
    bcac %>% 
      select(ID, SYMBOL, MASK, number_chrom, pos, P.bcac = P, z_score.bcac = z_score, oriented_beta.bcac = oriented_beta), 
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) %>%
  inner_join (
    jurgens %>% 
      select(ID, SYMBOL, MASK, number_chrom, pos, P.jurgens = P, z_score.jurgens = z_score, oriented_beta.jurgens = oriented_beta), 
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) 

weights <- combine %>% 
  filter(SYMBOL == "CHEK2", MASK == "PTV") %>% 
  mutate(
    b = z_score.bcac / z_score.mws,
    j = z_score.jurgens / z_score.mws,
    m = 1) %>%
  select(m, b, j) %>%
  as.list()

meta <- combine %>% 
  mutate(
    num = (z_score.mws * weights$m) + (z_score.bcac * weights$b) + (z_score.jurgens * weights$j),
    denom = sqrt(weights$m^2 + weights$b^2 + weights$j^2),
    z_score.meta = num/denom,
    P.meta = 2*pnorm(q=-abs(z_score.meta))
    ) %>%
  select(-starts_with("weight_"), -num, -denom) %>%
  pivot_longer(cols=c(starts_with("P."), starts_with("z_score."), starts_with("oriented_beta.")), 
               names_sep="\\.", names_to=c(".value", "Study"))

meta %>% arrange(P) %>% filter(Study=="meta", MASK=="PTV") %>% head(n=10)

meta_ids <- meta %>% pull(ID) %>% unique()

data.table::fwrite(meta, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta.csv")

```

```{r}

meta <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta.csv")

meta_ids <- meta %>% pull(ID) %>% unique()

meta_sig <- meta %>% 
  group_by(MASK) %>%
  mutate(threshold = 0.05/n_distinct(SYMBOL)) %>%
  select(MASK, threshold) %>% unique()

```


# White only

```{r, eval=FALSE}

regenie <- vector("list", 3)
i <- 1
for(file in c("rare", "cadd", "helix")) {
  filename = paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/regenie/step2_gene/binary/breast_cancer_w_",
                    file, "_breast_cancer_w.regenie")
  data <- data.table::fread(filename)
  
  # Rearrange
  data <- data %>% 
    separate_wider_delim(cols=ID, delim=".", names=c("SYMBOL", "MASK", "FREQ"), too_many="merge", too_few="align_start") %>%
    mutate(oriented_beta = BETA,
           P = 10^(-LOG10P),
           number_chrom=as.numeric(CHROM),
           pos = GENPOS)
  
  if(file == "rare") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV", MASK == "M2" ~ "Rare", TRUE ~ "Uhoh"))
    } else if(file == "cadd") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.CADD", MASK == "M2" ~ "CADD", TRUE ~ "Uhoh"))
    } else if(file == "helix") {
      data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.Helix", MASK == "M2" ~ "Helix", TRUE ~ "Uhoh"))
    }
  data <- data %>% mutate(ID = paste0(SYMBOL, "_", MASK, "_", FREQ, "_", TEST))
  regenie[[i]] <- data
  i <- i + 1
}

mws_w <- dplyr::bind_rows(regenie) %>% 
  filter(
    FREQ == "all",
    TEST == "ADD",
    MASK %in% c("PTV", "Rare", "PTV.CADD", "PTV.Helix")
  ) %>%
  mutate(
    ID = paste0(SYMBOL, "_", MASK),
    z_score = BETA/SE,
    MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix"))
    )
  
rm(regenie)

data.table::fwrite(mws_w, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/mws_w.csv")

```

```{r}

mws_w <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/mws_w.csv")

```

```{r}

bcac <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/bcac.csv")

```

```{r, eval=FALSE}

# dbname <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/anno_full.db"
# con <- DBI::dbConnect(duckdb::duckdb(), dbname)
# 
# colnames <- tbl(con, "annotation") %>% head() %>% collect()
# 
# genes <- tbl(con, "annotation") %>%
#   filter(Feature_type == "Transcript",
#          (is.na(MANE_SELECT) & CANONICAL == "YES") | !is.na(MANE_SELECT),
#          BIOTYPE == "protein_coding"
#          ) %>%
#   select(Gene, SYMBOL) %>%
#   distinct() %>%
#   collect()
# 
# DBI::dbDisconnect(con, shutdown=TRUE)
# 
# Jurgens_AoU_MGB <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/JurgensWang2024_all_files/All_EU_raw/AoU_250k_MGB_53k__10CarriersPerStudy_20Overall_cauchy_harmonizedEstimates.tsv.gz") %>%
#   separate_wider_delim(total, delim="__", names=c("Pheno", "Gene")) %>%
#   filter(Pheno=="Breast_cancer") %>%
#   inner_join(genes, by="Gene")
# 
# data.table::fwrite(Jurgens_AoU_MGB, file="/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/AoU_MGB_EU_breast_cancer.tsv")

Jurgens_AoU_MGB <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jurgens2024/AoU_MGB_EU_breast_cancer.tsv") %>% 
  select(
    Gene, SYMBOL, cases, controls, P_cauchy, P_cauchy_v2,
    ends_with("hclof_noflag_POPMAX0.001"), # Meta with PTV
    ends_with("missense0.2_POPMAX0.00001"), # Meta with rare missense
    ends_with("hclof_noflag_missense0.8_POPMAX0.001"), ends_with("hclof_noflag_missense0.5_POPMAX0.001") # Meta with PTV_CADD and PTV_Helix
    ) %>%
  rename(popcases = cases, popcontrols=controls) %>%
  pivot_longer(cols=contains("POPMAX"), names_to=c(".value", "raw_mask"),  names_sep="__") %>%
  inner_join(mws %>% select(number_chrom, pos, SYMBOL) %>% distinct(), by="SYMBOL") %>%
  filter(!is.na(P_Meta)) %>%
  mutate(
    P = P_Meta,
    old_z = qnorm(p=P_Meta/2, lower.tail=FALSE),
    z_score = Est_Meta/Est.SE_Meta,
    oriented_beta = Est_Meta,
    ID = SYMBOL
    ) %>%
  filter(!is.na(z_score))

jurgens_w <- Jurgens_AoU_MGB %>%
  filter(raw_mask != "hclof_noflag_missense0.5_POPMAX0.001") %>%
  mutate(MASK = dplyr::case_when(
    raw_mask == "hclof_noflag_POPMAX0.001" ~ "PTV",
    raw_mask == "missense0.2_POPMAX0.00001" ~ "Rare",
    raw_mask == "hclof_noflag_missense0.8_POPMAX0.001" ~ "PTV.CADD"
    )
  ) %>%
  rbind(
    Jurgens_AoU_MGB %>% 
      filter(raw_mask == "hclof_noflag_missense0.8_POPMAX0.001") %>% 
      mutate(MASK = "PTV.Helix")
  ) %>%
  mutate(ID = paste0(SYMBOL, "_", MASK))

data.table::fwrite(jurgens_w, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/jurgens_w.csv")

```

```{r}

jurgens_w <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/jurgens_w.csv")

```

## Meta-analysis of MWS/BCAC/UKB/AoU/MGB

```{r, eval=FALSE}

combine <- mws_w %>% 
  select(ID, SYMBOL, MASK, number_chrom, pos, P.mws = P, z_score.mws = z_score, oriented_beta.mws = oriented_beta) %>%
  inner_join(
    bcac %>%
      select(ID, SYMBOL, MASK, number_chrom, pos, P.bcac = P, z_score.bcac = z_score, oriented_beta.bcac = oriented_beta),
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) %>%
  inner_join (
    jurgens_w %>% 
      select(ID, SYMBOL, MASK, number_chrom, pos, P.jurgens = P, z_score.jurgens = z_score, oriented_beta.jurgens = oriented_beta), 
    by=c("ID", "SYMBOL", "MASK", "number_chrom", "pos")
    ) 

weights <- combine %>% 
  filter(SYMBOL == "CHEK2", MASK == "PTV") %>% 
  mutate(
    b = z_score.bcac / z_score.mws,
    j = z_score.jurgens / z_score.mws,
    m = 1) %>%
  select(m, j, b) %>% 
  as.list()

meta_w <- combine %>% 
  mutate(
    num = (z_score.mws * weights$m) + (z_score.jurgens * weights$j) + (z_score.bcac * weights$b),
    denom = sqrt(weights$m^2 + weights$j^2 + weights$b^2), 
    z_score.meta = num/denom,
    P.meta = 2*pnorm(q=-abs(z_score.meta))
    ) %>%
  select(-starts_with("weight_"), -num, -denom) %>%
  pivot_longer(cols=c(starts_with("P."), starts_with("z_score."), starts_with("oriented_beta.")), 
               names_sep="\\.", names_to=c(".value", "Study"))

data.table::fwrite(meta_w, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta_w.csv")

```

```{r}

meta_w %>% 
  filter(Study == "meta",
         MASK == "PTV",
         P < 1e-5
         )%>%
  arrange(MASK, P) %>% 
  mutate(
    Zscore = formatC(z_score, digits=2, format="f"),
    Pval = formatC(P, digits=2, format="e")
    ) %>%
  select(MASK, SYMBOL, Zscore, Pval)

```

```{r}

meta_w <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta_w.csv")

```

# Sensitivity and supplementary analyses

```{r, eval=FALSE}

read_burden <- function(pheno) {
  regenie <- vector("list", 3)
  i <- 1
  for(file in c("rare", "cadd", "helix")) {
    filename = paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/regenie/step2_gene/binary/",
    pheno, "_", file, "_", pheno, ".regenie")
    data <- data.table::fread(filename)
    
    # Rearrange
    data <- data %>% 
      separate_wider_delim(cols=ID, delim=".", names=c("SYMBOL", "MASK", "FREQ"), too_many="merge", too_few="align_start") %>%
      mutate(oriented_beta = BETA,
             P = 10^(-LOG10P),
             number_chrom=as.character(CHROM),
             pos = GENPOS)
    
    if(file == "rare") {
        data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV", MASK == "M2" ~ "Rare", TRUE ~ "Uhoh"))
      } else if(file == "cadd") {
        data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.CADD", MASK == "M2" ~ "CADD", TRUE ~ "Uhoh"))
      } else if(file == "helix") {
        data <- data %>% mutate(MASK = dplyr::case_when(MASK == "M1" ~ "PTV.Helix", MASK == "M2" ~ "Helix", TRUE ~ "Uhoh"))
      }
    data <- data %>% 
      filter(
        FREQ == "all",
        TEST == "ADD",
        MASK %in% c("PTV", "Rare", "PTV.CADD", "PTV.Helix")
      ) %>% 
      mutate(
        ID = paste0(SYMBOL, "_", MASK),
        z_score = BETA/SE
        )
    
    regenie[[i]] <- data
    i <- i + 1
  }
  
  mws <- dplyr::bind_rows(regenie) %>% mutate(MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix")))
    
  rm(regenie)
  return(mws)
}

breast_cancer_burden <- read_burden("breast_cancer")
breast_cancer_w_burden <- read_burden("breast_cancer_w")
breast_cancer_insitu_burden <- read_burden("breast_cancer_insitu")
bc_noCVD_burden <- read_burden("bc_noCVD_controls")
bc_before_55_burden <- read_burden("bc_before_55")
bc_after_55_burden <- read_burden("bc_after_55")
bc_screen_burden <- read_burden("bc_screen")
bc_interval_burden <- read_burden("bc_interval")
bc_withFH_strat_burden <- read_burden("bc_withFH_strat")
bc_noFH_strat_burden <- read_burden("bc_noFH_strat")

all_burden <- bind_rows(
  list(
    mws_breast_cancer=breast_cancer_burden, 
    overall_breast_cancer=breast_cancer_burden,
    breast_cancer_finn=breast_cancer_burden %>% filter(MASK == "PTV"),
    breast_cancer_w=breast_cancer_w_burden, 
    breast_cancer_insitu=breast_cancer_insitu_burden, 
    bc_noCVD=bc_noCVD_burden,
    bc_before_55=bc_before_55_burden,
    bc_after_55=bc_after_55_burden,
    bc_screen=bc_screen_burden, 
    bc_interval=bc_interval_burden,
    bc_withFH_strat=bc_withFH_strat_burden,
    bc_noFH_strat=bc_noFH_strat_burden
    ), 
  .id="Analysis"
  )

data.table::fwrite(all_burden, "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/supplementary.csv")

```

```{r}

all_burden <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/supplementary.csv")

```


# Case/control carrier counts by burden mask (not individual variants)

```{r, eval=FALSE}

carrier_count <- function(phenotype, burden_results) {
  masks <- c("PTV", "Rare", "PTV.CADD", "PTV.Helix")
  pheno <-data.table::fread(paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/exwas/phenofiles/", phenotype, "_pheno.txt")) %>% 
    select(-FID)
  
  variant_list  <- vector("list", 4)
  names(variant_list) <- masks
  
  for (mask in masks) {
    name <- dplyr::case_when(
      mask %in% c("PTV", "Rare") ~ "rare",
      mask == "PTV.CADD" ~ "cadd",
      mask == "PTV.Helix" ~ "helix",
      TRUE ~ "Oops"
    )
    mask_file <- paste0("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/regenie/step2_gene/binary/", phenotype, "_", name, "_masks")

    file_masks <- c("PTV" = "M1", "Rare" = "M2", "PTV.CADD" = "M1", "PTV.Helix" = "M1")
    
    sig_genes <- burden_results %>%
      filter(MASK == mask, P < 0.001) %>%
      pull(SYMBOL) %>%
      unique()
    
    genes <- data.table::fread(paste0(mask_file, ".bim"), col.names=c("CHR", "ID", "POS", "BP", "A1", "REF")) %>%
      separate_wider_delim(cols=ID, delim=".", names=c("SYMBOL", "MASK", "FREQ"), too_many="merge", cols_remove=FALSE) %>%
      filter(
        FREQ == "all", 
        SYMBOL %in% sig_genes, 
        MASK == file_masks[[mask]])
    
    if(nrow(genes) > 0) {
      raw_genes <- extract_variants_bfile(
        genes %>% pull(ID), 
        bfile = mask_file,
        plink_bin = config$software$plink2
        )
      
      variants <- as.data.frame(raw_genes) %>%
        select(-FID, -PAT, -MAT, -SEX, -PHENOTYPE) %>%
        mutate(across(contains("ref"), ~2-as.numeric(.x))) %>%
        pivot_longer(cols=contains("ref"), names_to="ID", values_to="carrier") %>%
        separate_wider_delim(cols=ID, delim=".", names=c("SYMBOL"), too_many="drop")
  
      variant_list[[mask]] <- variants
    } else {
      variant_list[[mask]] <- NULL
    }
    
  }
  
  rm(raw_genes)
  rm(variants)
  
  all_burden <- bind_rows(variant_list, .id="MASK") %>% 
    inner_join(pheno, by="IID") %>%
    filter(!is.na(carrier)) %>%
    mutate(
      MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix")),
      status = factor(.data[[phenotype]], levels=c(1, 2), labels=c("control", "case"))
      ) %>%
    select(IID, SYMBOL, MASK, status, carrier)
  
  rm(variant_list)
  
  return(all_burden)
}

mws_breast_cancer_counts <- carrier_count("breast_cancer_insitu", burden_results = breast_cancer_burden)

meta <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta.csv")
overall_breast_cancer_counts <- carrier_count("breast_cancer", burden_results = meta %>% filter(Study == "meta"))

meta_finn <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta_finn.csv")
breast_cancer_finn_counts <- carrier_count("breast_cancer", burden_results = meta_finn %>% filter(Study == "meta"))

meta_w <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/meta/data/meta_w.csv")
breast_cancer_w_counts <- carrier_count("breast_cancer_w", burden_results = meta_w %>% filter(Study == "meta"))

breast_cancer_insitu_counts <- carrier_count("breast_cancer_insitu", burden_results = breast_cancer_insitu_burden)
bc_noCVD_counts <- carrier_count("bc_noCVD_controls", burden_results = bc_noCVD_burden)
bc_before_55_counts <- carrier_count("bc_before_55", burden_results = rbind(bc_before_55_burden, bc_after_55_burden))
bc_after_55_counts <- carrier_count("bc_after_55", burden_results = rbind(bc_before_55_burden, bc_after_55_burden))
bc_screen_counts <- carrier_count("bc_screen", burden_results = rbind(bc_screen_burden, bc_interval_burden))
bc_interval_counts <- carrier_count("bc_interval", burden_results = rbind(bc_screen_burden, bc_interval_burden))
bc_withFH_strat_counts <- carrier_count("bc_withFH_strat", burden_results = rbind(bc_withFH_strat_burden, bc_noFH_strat_burden))
bc_noFH_strat_counts <- carrier_count("bc_noFH_strat", burden_results = rbind(bc_withFH_strat_burden, bc_noFH_strat_burden))

all_counts <- bind_rows(
  list(
    mws_breast_cancer=mws_breast_cancer_counts,
    overall_breast_cancer=overall_breast_cancer_counts,
    breast_cancer_finn=breast_cancer_finn_counts,
    breast_cancer_w=breast_cancer_w_counts, 
    breast_cancer_insitu=breast_cancer_insitu_counts,
    bc_noCVD=bc_noCVD_counts,
    bc_before_55=bc_before_55_counts, 
    bc_after_55=bc_after_55_counts,
    bc_screen=bc_screen_counts, 
    bc_interval=bc_interval_counts,
    bc_withFH_strat=bc_withFH_strat_counts,
    bc_noFH_strat=bc_noFH_strat_counts
    ), 
  .id="Analysis")

carrier_counts <- all_counts %>% 
  group_by(Analysis, MASK, SYMBOL) %>% 
  summarise(
    control_noncarriers = sum(status == "control" & carrier == 0),
    control_carriers = sum(status == "control" & carrier != 0), 
    case_noncarriers = sum(status == "case" & carrier == 0),
    case_carriers = sum(status == "case" & carrier != 0)
    ) %>% 
  arrange(Analysis, MASK, SYMBOL)

carrier_counts

```

## Tables

```{r}

format_pct <- function(pct) {
  if(pct < 0.01) {
    pretty_pct <- "<0.01"
  } else {
    pretty_pct <- formatC(pct, digits=2, format="f")
  }
  return(pretty_pct)
}

all_table <- all_burden %>% 
  select(Analysis, CHROM, SYMBOL, MASK, BETA, SE, P, z_score) %>%
  mutate(
    OR = formatC(exp(BETA), digits=2, format="f"),
    LCI = formatC(exp(BETA - 1.96*SE), digits=2, format="f"),
    UCI = formatC(exp(BETA + 1.96*SE), digits=2, format="f"),
    OR_CI = paste0(OR, "\n(", LCI, ", ", UCI, ")"),
    # Zscore = formatC(z_score, digits=2, format="f"),
    Pval = formatC(P, digits=2, format="e"),
    MASK = factor(MASK, levels=c("PTV", "Rare", "PTV.CADD", "PTV.Helix"))
    ) %>%
  inner_join(carrier_counts, by=c("Analysis", "MASK", "SYMBOL")) %>%
  mutate(
    control_pct = 100 * control_carriers / (control_carriers + control_noncarriers),
    case_pct = 100 * case_carriers / (case_carriers + case_noncarriers),
    control_pct = map(control_pct, format_pct),
    case_pct = map(case_pct, format_pct),
    control_carriers = paste0(control_carriers, "\n(", control_pct, "%)"),
    case_carriers = paste0(case_carriers, "\n(", case_pct, "%)")
    ) %>%
  arrange(Analysis, MASK, P) %>%
  select(Analysis, MASK, SYMBOL, control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval, P)

```

```{r}

############
# MWS only #
############

mws_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis == "mws_breast_cancer", P < 0.001) %>%
  arrange(MASK, P) %>%
  select(MASK, SYMBOL, OR_CI, Pval)
  
list_of_mws <- list(mws=mws_analyses)

#################
# Main analyses #
#################

main_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis == "overall_breast_cancer") %>%
  select(-P) %>%
  inner_join(
    meta %>% 
      filter(Study == "meta", P < 0.001) %>% 
      mutate(P.meta = formatC(P, digits=2, format="e"),
             Z.meta = formatC(z_score, digits=2, format="f")) %>%
      select(MASK, SYMBOL, P.m=P, Z.meta, P.meta),
    by=c("MASK", "SYMBOL")
  ) %>%
  arrange(MASK, P.m) %>%
  select(Analysis, MASK, SYMBOL, control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval, Z.meta, P.meta)
  
list_of_main <- split(main_analyses, main_analyses$MASK) %>%
  lapply(., function(x) {x %>% select(-Analysis, -MASK)})

####################
# PTV with FinnGen #
####################

finn_analysis <- all_table %>%
  ungroup() %>%
  filter(Analysis == "breast_cancer_finn") %>%
  select(-P) %>%
  inner_join(
    meta_finn %>% 
      filter(Study == "meta", P < 0.001) %>% 
      mutate(P.meta = formatC(P, digits=2, format="e"),
             Z.meta = formatC(z_score, digits=2, format="f")) %>%
      select(MASK, SYMBOL, P.m=P, Z.meta, P.meta),
    by=c("MASK", "SYMBOL")
  ) %>%
  arrange(MASK, P.m) %>%
  mutate(MASK = paste0(Analysis, ".", MASK)) %>%
  select(Analysis, MASK, SYMBOL, control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval, Z.meta, P.meta)
  
list_of_finn <- split(finn_analysis, finn_analysis$MASK, drop=TRUE) %>%
  lapply(., function(x) {x %>% select(-Analysis, -MASK)})

###############################
# Dichotomous supplementaries #
###############################

agestrat_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis %in% c("bc_before_55", "bc_after_55")) %>%
  select(-P) %>%
  arrange(desc(Analysis)) %>%
  pivot_wider(
    names_from=Analysis, 
    values_from=c(control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval)) %>%
  select(MASK, SYMBOL, ends_with("before_55"), ends_with("after_55")) %>%
  mutate(Analysis = "age_strat")

list_of_agestrat <- split(agestrat_analyses, 
                             list(agestrat_analyses$Analysis, agestrat_analyses$MASK), 
                             drop=TRUE, lex.order = TRUE) %>%
  lapply(., function(x) {x %>% select(-Analysis, -MASK)})

screenstrat_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis %in% c("bc_screen", "bc_interval")) %>%
  select(-P) %>%
  arrange(desc(Analysis)) %>%
  pivot_wider(
    names_from=Analysis, 
    values_from=c(control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval)) %>%
  select(MASK, SYMBOL, ends_with("bc_screen"), ends_with("bc_interval")) %>%
  mutate(Analysis = "screen_strat")

list_of_screenstrat <- split(screenstrat_analyses, 
                             list(screenstrat_analyses$Analysis, screenstrat_analyses$MASK), 
                             drop=TRUE, lex.order = TRUE) %>%
  lapply(., function(x) {x %>% select(-Analysis, -MASK)})

FHstrat_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis %in% c("bc_withFH_strat", "bc_noFH_strat")) %>%
  select(-P) %>%
    arrange(Analysis) %>%
  pivot_wider(
    names_from=Analysis,
    values_from=c(control_noncarriers, control_carriers, case_noncarriers, case_carriers, OR_CI, Pval)) %>%
  select(MASK, SYMBOL, ends_with("bc_noFH_strat"), ends_with("bc_withFH_strat")) %>%
  mutate(Analysis = "FH_strat")

list_of_FHstrat <- split(FHstrat_analyses,
                             list(FHstrat_analyses$Analysis, FHstrat_analyses$MASK),
                             drop=TRUE, lex.order = TRUE) %>%
  lapply(., function(x) {x %>% select(-Analysis, -MASK)})

##############
# White meta #
##############

white_analyses <- meta %>% 
  filter(Study == "meta") %>% 
  mutate(P.meta = formatC(P, digits=2, format="e"),
         Z.meta = formatC(z_score, digits=2, format="f")) %>%
  select(MASK, SYMBOL, P.m=P, Z.meta, P.meta) %>%
  full_join(
    meta_w %>% 
      filter(Study == "meta") %>% 
      mutate(P.eur = formatC(P, digits=2, format="e"),
             Z.eur = formatC(z_score, digits=2, format="f")) %>%
      select(MASK, SYMBOL, P.w=P, P.eur, Z.eur),
    by=c("MASK", "SYMBOL")
  ) %>%
  filter(P.m < 0.001 | P.w < 0.001) %>%
  arrange(MASK, P.m, P.w) %>%
  select(MASK, SYMBOL, Z.meta, P.meta, Z.eur, P.eur)
  
list_of_white <- list(Euro = white_analyses)

#######################
# Other supplementary #
#######################

supplementary_analyses <- all_table %>%
  ungroup() %>%
  filter(Analysis %in% c("breast_cancer_insitu", "bc_noCVD")) %>%
  select(Analysis, MASK, SYMBOL, OR_CI, Pval)

list_of_supplementary <- split(supplementary_analyses,
                             list(supplementary_analyses$Analysis),
                             drop=TRUE, lex.order = TRUE) %>%
  lapply(., function(x) {x %>% select(-Analysis)})

# Combine

list_of_datasets <- c(list_of_mws, list_of_main, list_of_finn, 
                      list_of_agestrat, list_of_screenstrat, list_of_FHstrat,
                      list_of_white, list_of_supplementary)

openxlsx::write.xlsx(list_of_datasets, file = "/home/jenniferco/MWS/Manuscript/Tables/SupplementaryUnformatted.xlsx", vdpi=300)

```

