---
title: "RunRegenie"
author: "Jennifer Collister"
date: "10/25/2024"
output: html_document
---

```{r setup, include=FALSE}

library(yaml)
library(here)
library(data.table)
library(R.utils)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(qqman)
library(ggpubr)
library(janitor)
library(ggrepel)
library(forcats)
library(DBI)
library(duckdb)
library(dbplyr)
library(tictoc)
library(glue)

knitr::opts_chunk$set(echo = TRUE)
# Set the root directory to the project directory (otherwise knitr works in the directory of the Rmd file by default)
knitr::opts_knit$set(root.dir = here::here())

# Load the project config file for filepaths etc
config = read_yaml(here::here("./config.yml"))

# Specify a safe tmp dir
unixtools::set.tempdir(config$tmp)

# Source shared functions
source(config$scripts$shared)

```

# Run Regenie

Universal parameters

```{r}

outpath <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/regenie"

universal <- list(
  regenie = "/home/jenniferco/software/regenie_v3.4.1.gz_x86_64_Centos7_mkl",
  bedpath_gxs = "/shared/MWS_Regeneron_Data/MWS_regeneron/pVCF/GxS_PLINK/OXFORD-MWS_Freeze_One.GxS",
  bedpath_exome = "/shared/MWS_Regeneron_Data/MWS_regeneron/pVCF/exome_PLINK/REGENIE_FILES/OXFORD-MWS_Freeze_One.norm",
  pruned = "/shared/MWS_Regeneron_Data/MWS_regeneron/REGENIE_FILES/ld_pruned.prune.in",
  phenopath = "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/exwas/phenofiles",
  covarfile = "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/exwas/covarfile/covars.txt",
  mask_path = "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/summaryQC/backman2021/masks",
  outpath = outpath,
  step1_out = file.path(outpath, "step1"),
  step2_exwas = file.path(outpath, "step2_exwas"),
  step2_gene = file.path(outpath, "step2_gene")
)

```

Trait specific parameters

```{r}

binary_param <- list(
  pheno_cols = c(
    "breast_cancer",
    "breast_cancer_w",
    "breast_cancer_insitu",
    "bc_noCVD_controls",
    "bc_before_55",
    "bc_after_55",
    "bc_screen",
    "bc_interval",
    "bc_withFH_strat",
    "bc_noFH_strat"
    ), 
  covar_cols = paste(c(
    "Age",
    "pc1", "pc2", "pc3", "pc4", "pc5",
    "pc6", "pc7", "pc8", "pc9", "pc10"
  ), collapse=","),
  trait = "--bt --cc12",
  firth = "--firth --approx",
  out = "binary"
)

continuous_param <- list(
  pheno_cols = c(
    "age_breast_cancer"
    ), 
  covar_cols = paste(c(
    "pc1", "pc2", "pc3", "pc4", "pc5",
    "pc6", "pc7", "pc8", "pc9", "pc10"
  ), collapse=","),
  trait = "--qt",
  firth = "",
  out = "continuous"
)

```

Run using screen to launch detached sessions from within R.

`screen -dm -S [name] [cmd]`

* `-d` = detached
* `-m` = completely detached
* `-S` = name

Use `ionice -c2 -n7 nice` as requested by Dave to play nice on Morpheus.

## Step 1

```{r}

command_list <- list()

for (param_set in list(binary_param, continuous_param)) {
  print(param_set[['out']])
  
  for(pheno in param_set[["pheno_cols"]]) {
    print(pheno)
    pheno_file <- file.path(universal$phenopath, paste0(pheno, "_pheno.txt"))
    outpath <- file.path(universal[['step1_out']], param_set[['out']], pheno)
    
    regenie_cmd <- glue(
      "ionice -c2 -n7 nice",
      "{universal[['regenie']]}",
      "--step 1",
      "--bed {universal[['bedpath_gxs']]}",
      "--extract {universal[['pruned']]}",
      "--phenoFile {pheno_file}",
      "--phenoColList {pheno}",
      "--covarFile {universal[['covarfile']]}",
      "--covarColList {param_set[['covar_cols']]}",
      "{param_set[['trait']]}",
      "--bsize 1000 --threads 16",
      "--out {outpath}",
      .sep = " "
    )
    command_list[[pheno]] <- regenie_cmd
  }
}

run <- paste0("screen -dmS step1 ionice -c2 -n7 nice bash -c '", paste(command_list, collapse=" && "), "'")
run
system(run)

```

(Running all of these at once without `ionice` kinda tanked Morpheus, they were all nomming too much CPU)

## Step 2 ExWAS

```{r}

command_list <- list()

for (param_set in list(binary_param, continuous_param)) {
  print(param_set[['out']])
  
  for(pheno in param_set[["pheno_cols"]]) {
    print(pheno)
    pheno_file <- file.path(universal$phenopath, paste0(pheno, "_pheno.txt"))
    outpath <- file.path(universal[['step2_exwas']], param_set[['out']], pheno)
    
    regenie_cmd <- glue(
      "{universal[['regenie']]}",
      "--step 2",
      "--bed {universal[['bedpath_exome']]}",
      "--phenoFile {pheno_file}",
      "--phenoColList {pheno}",
      "--covarFile {universal[['covarfile']]}",
      "--covarColList {param_set[['covar_cols']]}",
      "--pred {file.path(universal[['step1_out']], param_set[['out']], pheno)}_pred.list",
      "{param_set[['trait']]} {param_set[['firth']]} --pThresh 0.01 --af-cc",
      "--bsize 400 --threads 16",
      "--out {outpath}",
      "--chrList 1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23",
      .sep = " "
    )
    command_list[[pheno]] <- regenie_cmd
  }
}
run <- paste0("screen -dmS step2_exwas ionice -c2 -n7 nice bash -c '", paste(command_list, collapse=" && "), "'")
run
system(run)

```

## Step 2 Gene burden

```{r}

setfile <- "BCAC_set.txt"

command_list <- list(rare=list(), cadd=list(), helix=list())

for (param_set in list(binary_param, continuous_param)) {
  print(param_set[['out']])
  
  for(pheno in param_set[["pheno_cols"]]) {
    print(pheno)
    pheno_file <- file.path(universal$phenopath, paste0(pheno, "_pheno.txt"))
    outpath <- file.path(universal[['step2_gene']], param_set[['out']])
    
    for (mask in c("rare", "cadd", "helix")) {
      print(mask)
      annofile <- paste0("BCAC_", mask, "_missense_anno.txt")
      maskfile <- paste0("BCAC_", mask, "_missense_mask.txt")
      
      outfile <- paste0(pheno, "_", mask)
    
      regenie_cmd <- glue(
        "ionice -c2 -n7 nice",
        "{universal[['regenie']]}",
        "--step 2",
        "--bed {universal[['bedpath_exome']]}",
        "--phenoFile {pheno_file}",
        "--phenoColList {pheno}",
        "--covarFile {universal[['covarfile']]}",
        "--covarColList {param_set[['covar_cols']]}",
        "--pred {file.path(universal[['step1_out']], param_set[['out']], pheno)}_pred.list",
        "{param_set[['trait']]} {param_set[['firth']]}",
        "--af-cc --write-mask --write-mask-snplist",
        "--bsize 200 --threads 16",
        "--anno-file {file.path(universal[['mask_path']], annofile)}",
        "--set-list {file.path(universal[['mask_path']], setfile)}",
        "--mask-def {file.path(universal[['mask_path']], maskfile)}",
        "--out {file.path(outpath, outfile)}",
        "--aaf-bins 0.001,0.01,0.05,1",
        .sep = " "
      )
      command_list[[mask]][[pheno]] <- regenie_cmd
    }
  }
}
for (mask in c("rare", "cadd", "helix")) {
  run <- paste0("screen -dmS step2_gene_p_", mask, " ionice -c2 -n7 nice bash -c '", paste(command_list[[mask]], collapse=" && "), "'")
  system(run)
}

```

