---
title: "BCAC_Annotation"
author: "Jennifer Collister"
date: "22/08/2024"
output: html_document
---

```{r setup, include=FALSE}

library(yaml)
library(here)
library(data.table)
library(R.utils)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(qqman)
library(ggpubr)
library(janitor)
library(ggrepel)
library(forcats)
library(DBI)
library(duckdb)
library(dbplyr)
library(tictoc)
library(jsonlite)
library(httr2)

knitr::opts_chunk$set(echo = TRUE)
# Set the root directory to the project directory (otherwise knitr works in the directory of the Rmd file by default)
knitr::opts_knit$set(root.dir = here::here())

# Load the project config file for filepaths etc
config = read_yaml(here::here("./config.yml"))

# Specify a safe tmp dir
unixtools::set.tempdir(config$tmp)

# Source shared functions
source(config$scripts$shared)

```

# Get annotations from Helix

[Helix scores](https://helixlabs.ai/blog/helix_whitepaper/)

## Get variants in format required to annotate

Helix is available for missense variants only, so we'll filter to missense variants

```{r, eval=FALSE}

dbname <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/anno_full.db"
con <- DBI::dbConnect(duckdb::duckdb(), dbname)

helix_all <- tbl(con, "annotation") %>%
  filter(Feature_type == "Transcript",
         (is.na(MANE_SELECT) & CANONICAL == "YES") | !is.na(MANE_SELECT),
         BIOTYPE == "protein_coding") %>% 
  select(Uploaded_variation, Consequence, Gene, Feature, Feature_type, HGVSc, HGVSp) %>%
  mutate(ID = str_replace_all(Uploaded_variation, pattern="_", replacement=":")) %>%
  collect()

DBI::dbDisconnect(con, shutdown=TRUE)

helix_missense <- helix_all %>%
  mutate(Consequence = str_split_i(Consequence, pattern=",", i=1)) %>%
  filter(Consequence %in% c("stop_lost", "start_lost", "transcript_amplification",
                            "inframe_insertion", "missense_variant", 
                            "protein_altering_variant", "splice_region_variant"))

head(helix_missense)

helix_vars <- helix_missense %>% 
  separate_wider_delim(HGVSp, delim=":p.", names=c("ENSP", "Aminos"), cols_remove=FALSE) %>%
  mutate(acid1 = substr(Aminos, 1, 3),
         residue = substr(Aminos, 4, nchar(Aminos)-3),
         acid2 = substr(Aminos, nchar(Aminos)-2, nchar(Aminos))) %>%
  arrange(Feature, residue, acid2)

head(helix_vars)

helix_uniques <- helix_vars %>% select(-ID, -HGVSc, -Uploaded_variation) %>% distinct()

data.table::fwrite(helix_uniques, file="/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/helix/MWS_variants.txt", 
                   col.names=TRUE)

```

## Annotate

Run Helix_MWS.R

# Combine with existing annotation

We need

* 1000 genomes phase 3 allele frequency
* sequence ontology variant consequences 
* exon/intron number
* CADD score
* Helix score

The masks we need to produce are: PTV, rare missense, deleterious rare missense (CADD > 20) and deleterious rare missense (Helix > 0.5)

The BCAC paper doesn't specify how they define PTV, so we will use the definition from [Palmer2022](https://www.nature.com/articles/s41588-022-01034-x) Supplementary Table 5:

> Transcript ablation, splice acceptor variant, splice donor variant, stop gained, frameshift variant.

They also define missense as:

> Stop lost, start lost, transcript amplification, inframe insertion, inframe deletion, missense variant, protein altering variant, splice region variant.


This seems to broadly agree with [Liu2023](https://www.nature.com/articles/s41588-023-01305-1#Sec2)

> (1) PTVs, defined as any mutation that introduced a stop codon, changed the frame of the open reading frame or introduced a change at a predicted splice donor or splice acceptor site

And [Zhao2024](https://www.nature.com/articles/s41588-024-01694-x#Sec9)

> Stop-gained, splice acceptor and splice donor variants were merged into a combined PTV category, while annotations for missense and synonymous variants were adopted directly from VEP

The [All by All](https://github.com/atgu/aou_gwas/blob/c27c06caa9614251b48bc17d1495138cb3d58d1d/utils/annotations.py#L5) tables appear to use the same definitions as the [Neale lab](https://github.com/Nealelab/ukb_common/blob/9ec920ce9e0c39c4b9baf896ae820c9e0d7f167a/utils/annotations.py) UKB annotations.

> PLOF_CSQS = ["transcript_ablation", "splice_acceptor_variant",
             "splice_donor_variant", "stop_gained", "frameshift_variant"]

> MISSENSE_CSQS = ["stop_lost", "start_lost", "transcript_amplification",
                 "inframe_insertion", "inframe_deletion", "missense_variant"]
                 
but this additionally uses Loftee to categorise the PTVs as pLoF or LC.

The [Neale lab paper](https://www.cell.com/cell-genomics/fulltext/S2666-979X(22)00110-0) describes this as

> Variants were grouped by Ensembl Gene ID and functional impact as follows:
>
> * pLoF: High-confidence LoF variants (as indicated by LOFTEE), including stop-gained, essential splice, and frameshift variants, filtered according to a set of first principles as described at https://github.com/konradjk/loftee
> * missense|LC: Missense variants are grouped with in-frame insertions and deletions, as well as low-confidence LoF variants (filtered out by LOFTEE). The latter have a frequency spectrum consistent with missense variation, and affect a set of amino acids in a similar fashion (e.g. a frameshift in the final exon).


```{r}

dbname <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/anno_full.db"
con <- DBI::dbConnect(duckdb::duckdb(), dbname)

colnames <- tbl(con, "annotation") %>% head() %>% collect()

anno <- tbl(con, "annotation") %>%
  filter(Feature_type == "Transcript",
         (is.na(MANE_SELECT) & CANONICAL == "YES") | !is.na(MANE_SELECT),
         BIOTYPE == "protein_coding") %>%
  select(
    Uploaded_variation, Location, Allele, AF,
    Gene, EXON, HGVSp,
    Feature, Feature_type, Consequence, 
    IMPACT, VARIANT_CLASS, SYMBOL, 
    CADD_phred, CADD_raw, CADD_raw_rankscore
  ) %>%
  group_by(Uploaded_variation) %>%
  slice_max(IMPACT, with_ties=FALSE) %>%
  ungroup() %>%
  mutate(ID = str_replace_all(Uploaded_variation, pattern="_", replacement=":")) %>%
  collect()

DBI::dbDisconnect(con, shutdown=TRUE)

list_exonic <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/exon_target_regions/exonic_vars.txt", 
                                 col.names=c("var")) %>% 
  pull(var)

exonic <- anno %>% 
  separate_wider_delim(Location, delim="-", names=c("chrpos", "pos2"), too_few="align_start", cols_remove=FALSE) %>%
  filter(chrpos %in% list_exonic) %>%
  separate_wider_delim(chrpos, delim=":", names=c("chr", "pos"), cols_remove=TRUE) %>%
  separate_wider_delim(EXON, delim="/", names=c("exon_n", "exon_total")) %>%
  mutate(
    chr = str_remove(chr, "chr"),
    pos = as.integer(pos),
    exon_n = as.numeric(exon_n),
    exon_total = as.numeric(exon_total)
  )

penultimate <- exonic %>% 
  filter(
    exon_n == exon_total - 1
  ) %>%
  select(chr, pos, SYMBOL) %>%
  group_by(SYMBOL) %>%
  slice_max(pos) %>%
  ungroup() %>%
  unique() %>%
  mutate(
    pos_50 = pos - 50,
    pos_list = mapply(seq, from=pos_50, to=pos, by=1, SIMPLIFY = FALSE)
  ) %>%
  select(chr, pos_list, SYMBOL) %>%
  tidyr::unnest(pos_list) %>%
  mutate(chrpos = paste0(chr, ":", pos_list)) %>%
  pull(chrpos)

freq <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/freq/all.afreq") 
helix <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/annotation/helix/MWS_anno.txt")

annotations <- exonic %>%
  filter(!is.na(SYMBOL)) %>%
  mutate(MAF_1kg = ifelse(AF < 0.5, AF, 1-AF)) %>% 
  dplyr::left_join(helix, join_by(Feature == transcript_id, HGVSp == HGVSp)) %>%
  dplyr::left_join(
    freq %>%
      mutate(MAF = ifelse(ALT_FREQS < 0.5, ALT_FREQS, 1 - ALT_FREQS)) %>%
      select(ID, MAF), 
    by="ID") %>%
  mutate(
    Consequence = str_split_i(Consequence, pattern=",", i=1),
    PTV = dplyr::case_when(
      ! Consequence %in% c("transcript_ablation", "splice_acceptor_variant", "splice_donor_variant", "stop_gained", "frameshift_variant") ~ "Not a PTV",
      exon_n == exon_total ~ "Last exon",
      Location %in% penultimate ~ "Last 50bp of penultimate exon",
      TRUE ~ "PTV"
    ),
    missense = Consequence %in% c("stop_lost", "start_lost", "transcript_amplification",
                                         "inframe_insertion", "missense_variant", "protein_altering_variant", "splice_region_variant"),
    # If variant not present in 1kg, that's fine
    rare_missense = missense & MAF < 0.001 & (MAF_1kg < 0.001 | is.na(MAF_1kg)),
    cadd_missense = rare_missense & CADD_phred > 20,
    helix_missense = rare_missense & deleterious > 0.5 & residue_type == acid1,
  )

# rm(helix)
# rm(anno)
# rm(freq)
# rm(list_exonic)
# rm(exonic)

head(annotations)

annotations %>% filter(pos == 99318162)

```

### Restrict to variants that meet Regeneron QC:

* Hardy–Weinberg equilibrium: HWE threshold of 10^-15
* Locus missingness: Locus missingness < 10%

```{r}

qc_vars <- data.table::fread("/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/summaryQC/backman2021/excludeIDs_ALL.txt") %>%
  mutate(ID = str_replace(str_replace_all(ID, ":", "_"), "_", ":")) %>%
  pull(ID)

qc_fail <- annotations %>% filter(Uploaded_variation %in% qc_vars)

qc_succeed <- annotations %>% filter(!Uploaded_variation %in% qc_vars)

```

Excluding ones that fail this QC removes 19,810 variants from the annotated exonic regions.

## Output in Regenie format

### Annotation file

```{r}

anno_path <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/summaryQC/backman2021/masks"

for(annotation in c("rare_missense", "cadd_missense", "helix_missense")) {
  anno_file <- file.path(anno_path, paste0("BCAC_", annotation, "_anno.txt"))
  
  data <- qc_succeed %>% 
    filter(PTV == "PTV") %>%
    mutate(anno = "PTV") %>%
    select(ID, SYMBOL, anno)
  
  data.table::fwrite(data, anno_file, append=FALSE, sep="\t", col.names=FALSE)
  
  data <- annotations %>% 
    filter(.data[[annotation]] == TRUE) %>%
    mutate(anno = annotation) %>%
    select(ID, SYMBOL, anno)
  
  data.table::fwrite(data, anno_file, append=TRUE, sep="\t", col.names=FALSE)
}

```

### Set file

```{r}

set_file <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/summaryQC/backman2021/masks/BCAC_set.txt"

set <- qc_succeed %>%
  select(GENE=SYMBOL, ID) %>%
  separate_wider_delim(cols=ID, delim=":", names=c("CHR", "POS", NA, NA), cols_remove=FALSE) %>%
  arrange(CHR, POS) %>%
  group_by(GENE) %>%
  mutate(CHR = as.numeric(CHR),
         varlist = paste0(ID, collapse=",")) %>%
  slice_min(POS, with_ties=FALSE) %>%
  select(GENE, CHR, POS, varlist) %>%
  filter(CHR != 24) %>%
  arrange(CHR, GENE)

head(set)

data.table::fwrite(set, set_file, sep="\t", col.names=FALSE)
  
```

### Masks file

```{r}

mask_path <- "/shared/MWS_Regeneron_Data/MWS_regeneron/Jennifer/summaryQC/backman2021/masks"

for(annotation in c("rare_missense", "cadd_missense", "helix_missense")) {
  mask_file <- file.path(mask_path, paste0("BCAC_", annotation, "_mask.txt"))
  
  if(annotation == "rare_missense") {
     mask <- data.frame(mask=c("M1", "M2"), anno=c("PTV", "rare_missense"))
  } else {
    mask <- data.frame(mask=c("M1", "M2"), anno=c(paste0("PTV,", annotation), annotation))
  }
  data.table::fwrite(mask, mask_file, sep="\t", col.names=FALSE)
}

```

Now move to /home/jenniferco/MWS/Scripts/Burden/RunRegenie.Rmd
